A Wait-free Queue with Poly-logarithmic Worst-caseStep Complexity

### 2022-11-25 podc_paper/podc2023sample.tex
add garbage_collect+podc_papaer

\documentclass[11pt,letterpaper]{article}
\renewcommand{\baselinestretch}{1.8}

\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{showkeys}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{relsize}
\usepackage{enumitem}
\usepackage{bold-extra}
\renewcommand*\contentsname{Table of Contents}

@@

\begin{document}

\begin{center}
\pagenumbering{roman}
\huge{\bf{A Wait-free Queue with Poly-logarithmic\\Worst-case Step Complexity}}

\vspace{2em}
\Large{\bf{Hossein Naderibeni}}

\vspace{3em}
\large A Thesis submitted to the Faculty of Graduate Studies \\ in Partial Fulfillment of the Requirements\\for the Degree of Master of Science

\vspace{2em}
Graduate Program in Computer Science,\\
York University\\
Toronto, Ontario

\vspace{2em}
November, 2022

\vspace{3em}
\textcopyright \; Hossein Naderibeni, 2022

\end{center}
\thispagestyle{empty}
\pagebreak

\normalsize
\addcontentsline{toc}{subsection}{Abstract}
\section*{Abstract}

In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem. We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.

\addcontentsline{toc}{subsection}{Acknowledgements}
\section*{Acknowledgements}
\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert. Without his guidance and support, this thesis probably would not have been completed. He helped me in every step of this thesis, taught me shared data structures, helped me find the problem, design and improve the algorithm, and write it with its complications because of lots of new notions in it. Thank you again for always pushing to make it better.

I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments but also for his academic advice and motivations.

Besides my supervisors, I am grateful to all members of my examining committee for their time and their helpful comments on my thesis.

Last but not least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study. I want to thank my brother for being there for me whenever I needed help.

I would like to express my gratitude to York University for the scholarship I was awarded in 2021-2022 academic year.

I dedicate this paper to the people in Iran and all the people who are fighting for their freedom.

\pagebreak
\addcontentsline{toc}{subsection}{Table of Contents}
\tableofcontents

\pagebreak
\addcontentsline{toc}{subsection}{List of Tables}
\listoftables

\pagebreak
\addcontentsline{toc}{subsection}{List of Figures}
\listoffigures
\newpage

\pagenumbering{arabic}
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same. In this run, $P$'s update is overwritten, which makes the queue inconsistent. One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

@@

The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response. Figure \ref{fig::goodexample} depicts an example of an execution on a linearizable  queue that is initially empty. The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation. Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}. The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/good}
  \caption[An example of a linearizable execution.]{\label{fig::goodexample}An example of a linearizable execution. Either \texttt{Enqueue(A)} or \texttt{Enqueue(B)} could take effect first since they are concurrent.}
\end{figure}

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption[An example of a non-linearizable execution.]{\label{fig::badexample}An example of an execution on an empty queue that is not linearizable. \texttt{Enqueue(A)} has terminated before \texttt{Enqueue(B)} is invoked and the \texttt{Dequeue} can occur before \texttt{Enqueue(A)} and return \nf{null} or can occur after \texttt{Enqueue(A)} and return \texttt{A}.}
\end{figure}

@@

\paragraph{Thesis outline} The rest of the thesis is organized as follows. Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation. Previous lock-free queues and their common problem are presented in Section 2.1. In Section 2.2, we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.

@@

\pagebreak
\section{Related Work}
In this section, we look at previous lock-free queues. The amortized step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the worst case, where all $p$ processes try to do an enqueue or a dequeue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

\subsection{List-based Queues}
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. A version of it is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}). A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue. To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the amortized number of steps $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.4]{pics/msqueue}
  \caption[MS-queue structure.]{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. In the first diagram, the first element has been dequeued. Red arrows show new pointers and gray dashed arrows show the old pointers.}
\end{figure}

@@

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption[Baskets queue.]{At some time, all operations in a basket ran concurrently, but only one succeeded to do \texttt{CAS}. To order the operations in a basket, processes perform another \nf{CAS}. The successful process will be the next one in the basket, and so on.}
\end{figure}

@@

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. The queue is stored in a shared array where head and tail pointers point to the current elements in the queue. When the array is full, an empty array is linked to the array and tail pointers are updated. A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array. Global tail and local tail pointers are similar (see~Figure~\ref{fig::sundell}). A process updates the position of the pointers after it does an operation. One process might go to sleep before setting the pointers, so the pointers might be behind their real places. They mention how to scan the arrays to update pointers while doing an operation. A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the amortized step (and \nf{CAS}) complexity remains $\Omega(p)$. Their design is lock-free but not wait-free.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/sundell}
    \caption[Global and local pointers in~\cite{DBLP:conf/opodis/GidenstamST10}.]{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
\end{figure}

@@

% TODO: page

@@

\newpage
\clearpage

\subsection{Example Execution}

Here we want to demonstrate how four processes help one another to  propagate their operations up to the root and then compute their responses.

Initially, each node in the tree contains one empty block. In Figure \ref{fig::exec1}, processes $P1,P2,P4$ have appended the blocks containing $ENQ_1,DEQ_2,DEQ_4$. Line \ref{appendLeaf} appends a block to the process's leaf.

Then, as shown in Figure \ref{fig::exec2}, $P2$ does a successful \nf{Refresh} on the node $n5$ that propagates both the operations pf $P1$ and $P2$. $P4$ does also a successful \nf{Refresh} on the node $n6$ and propagates $DEQ_4$. $P1$'s \nf{Refresh} on $n5$ is slower than $P2$'s and creates its new block after $P2$'s successful \nf{CAS} that propagated $ENQ_1$ and $ENQ_2$. So the block created by $P1$ is empty as there is nothing to propagate ($P1$'s Line \ref{readHead} is after $P2$'s Line \ref{cas}).

Since the block that $P1$ creates is empty, it returns \nf{true} in Line \ref{addOP} as shown in Figure \ref{fig::exec3}. $P3$ propagates $ENQ_3$ to $n6$ and $P4$ is very fast and it propagates $DEQ_4$ to the \nf{root} even before $P2$ puts its block into $n5$.

Then, as shown in Figure~\ref{fig::exec4}, $P4$ goes to sleep and $P2$ helps to increment \nf{root.head} after its failed \nf{CAS} on the root. After incrementing \nf{root.head}, $P3$ succeeds to do \nf{root.Refresh} with a single \nf{CAS} on the \nf{root} and puts operations from $P1,P2$ and $P3$ into the \nf{root}. $P1$ fails to do \nf{Refresh} on the \nf{root} two times, but it is guaranteed that its operation is propagated before its second failed \nf{CAS} on \nf{root.blocks}.

All the fields of selected blocks from Figure \ref{fig::exec4} are shown in Figure \ref{fig::exec5}. $P1$ and $P3$'s operations terminated after that they appended the \nf{Enqueue}s to the root. $P2$ needs to compute the index of the block that its \nf{Dequeue} is in. It computes the superblock of its leaf block, which is the second block in $n5$. Then, $P2$ computes the superblock of the block containing $ENQ_1,DEQ_2$. As we can see the \nf{super} field of  $ENQ_1,DEQ_2$ is 1 but the index of its superblock is 2. The queue size after the previous block is 0 so $DEQ_2$'s response is $ENQ_1$'s element. Then $P2$ tries to get $ENQ_1$'s element. $ENQ_1$ is in the left subblock of  block $ENQ_1,ENQ_3,DEQ_2$ because 1 is less than the \nf{sum\sub{enq-left}} field of the block containing $ENQ_1,ENQ_3,DEQ_2$. Similarly, $P2$ determines that $ENQ_1$ came from the left child of $n5$. $P_2$ therefore reaches the leaf $n1$ and returns $ENQ_1$'s element. $P4$ similarly finds the block in the \nf{root} that contains $DEQ_4$. Since it is the first operation in the block and the size of the queue after the previous (empty) block is 0, $DEQ_4$ returns \nf{null}.

\begin{figure}
  \center\includegraphics[width=4.8in]{thesis/pics/exec1.png}
  \caption{$P1,P2,P4$ append operations to their leaves. \label{fig::exec1}}
\end{figure}

\begin{figure}
  \center\includegraphics[width=4.8in]{thesis/pics/exec2.png}
  \caption[$P2$ helps $P1$. $P4$ is faster than $P3$.]{In the diagram, we use $n[i]$ as an abbreviation for $n$\nf{.blocks[$i$]}. $P2$ helps $P1$. $P4$ is faster than $P3$. \label{fig::exec2}}
\end{figure}

\begin{figure}
  \center\includegraphics[width=4.8in]{thesis/pics/exec3.png}
  \caption{$P4$ puts its block into the \nf{root} before $P2$ appends its block to $n5$.\label{fig::exec3}}
\end{figure}

\begin{figure}
  \center\includegraphics[width=4.8in]{thesis/pics/exec4.png}
  \caption[$P3$'s \nf{CAS} on \nf{root.blocks} succeeds.]{$P3$'s \nf{CAS} on \nf{root.blocks} succeeds. $P1$ and $P2$'s operations are done, each after two failed \nf{CAS} attempts on \nf{root.blocks}. \label{fig::exec4}}
\end{figure}

\begin{figure}
  \center\includegraphics[width=4.8in]{thesis/pics/exec5.png}
  \caption{Figure \ref{fig::exec4} with completed fields. \label{fig::exec5}}
\end{figure}

\clearpage

@@

\addcontentsline{toc}{section}{References}
\bibliography{main.bib} 
\bibliographystyle{acm}

\end{document}

>>>>

%%
%% This is file generated with the docstrip utility from the source file: 
%%
%% samples.dtx  (with options: `acmsmall-conf')
%% With some modifications for PODC 2023.
%% For the copyright see the source file.
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,nonacm,anonymous]{acmart}

\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{relsize}
\usepackage{enumitem}
\usepackage{bold-extra}
\renewcommand*\contentsname{Table of Contents}

@@

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2023}
\acmYear{2023}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[PODC 2023]{ACM Symposium on Principles of Distributed Computing}{June 19--23,
  2023}{Orlando, FL}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{A Wait-free Queue with Poly-logarithmic Worst-caseStep Complexity}
%\title{The Name of the Title Is Hope}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Eric Ruppert}
\affiliation{%
  \institution{York University}
  \streetaddress{P.O. Box 1212}
  \city{Toronto}
  \state{Ontario}
  \country{Canada}
  \postcode{43017-6221}
}
\authornote{Both authors contributed equally to this research.}
\email{ruppert@cse.yorku.ca}
\orcid{1234-5678-9012}

\author{Hossein Naderibeni}
\authornotemark[1]
\email{hnaderi268@gmail.com}
\affiliation{%
  \institution{York University}
  \streetaddress{P.O. Box 1212}
  \city{Toronto}
  \state{Ontario}
  \country{Canada}
  \postcode{43017-6221}
}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
In this work, we introduce a novel linearizable wait-free queue implementation.
Linearizability and lock-freedom are standard requirements for designing shared data structures. To the best of our knowledge, all of the existing linearizable lock-free queues in the literature have a common problem in their worst case, called the CAS Retry Problem. We show that our algorithm avoids this problem with the helping mechanism which we use and has a worst-case running time better than prior lock-free queues. The amortized number of steps for an \nf{Enqueue} or \nf{Dequeue} in our algorithm is $O(\log^2 p + \log q)$,  where $p$ is the number of processes and $q$ is the size of the queue when the operation is linearized.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{concurrent data structures, wait-free queues}

\received{20 February 2007}
\received[revised]{12 March 2009}
\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same. In this run, $P$'s update is overwritten, which makes the queue inconsistent. One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

@@

The standard safety condition is called \textit{linearizability}~\cite{10.1145/78969.78972}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response.

@@

The rest of the thesis is organized as follows. Chapter 2 gives an outline of the related work done in the area and the motivation for our implementation. Previous lock-free queues and their common problem are presented in Section 2.1. In Section 2.2, we mention some restricted lock-free queues. Section 2.3 talks about poly-logarithmic constructions of shared objects. Section 2.4 ends with a lower bound on the amortized time complexity of shared queues.

@@

\section{Related Work}
In this section, we look at previous lock-free queues. The amortized step complexity of all the lock-free queues includes an $\Omega(p)$ term that comes from the worst case, where all $p$ processes try to do an enqueue or a dequeue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

\subsection{List-based Queues}
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. A version of it is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly-linked list. A shared variable Head points to the first node in the linked list that has not been dequeued, and the Tail points to the last element in the queue. To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the amortized number of steps $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

@@

@@

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. The queue is stored in a shared array where head and tail pointers point to the current elements in the queue. When the array is full, an empty array is linked to the array and tail pointers are updated. A global head points to the array containing the first element in the queue, and each process has a local head index that points to the first element in that array. Global tail and local tail pointers are similar. A process updates the position of the pointers after it does an operation. One process might go to sleep before setting the pointers, so the pointers might be behind their real places. They mention how to scan the arrays to update pointers while doing an operation. A process writes an element in the location head by a \nf{CAS} instruction, so if $p$ processes try to enqueue simultaneously, the amortized step (and \nf{CAS}) complexity remains $\Omega(p)$. Their design is lock-free but not wait-free.

@@

@@

@@

\section{Template Overview}
As noted in the introduction, the ``\verb|acmart|'' document class can
be used to prepare many different kinds of documentation --- a
double-blind initial submission of a full-length technical paper and more --- all by
selecting the appropriate {\itshape template style} and {\itshape
  template parameters}.

This document will explain the major features of the document
class. For further information, the {\itshape \LaTeX\ User's Guide} is
available from
\url{https://www.acm.org/publications/proceedings-template}.

\subsection{Template Styles}

The primary parameter given to the ``\verb|acmart|'' document class is
the {\itshape template style} which corresponds to the kind of publication
or SIG publishing the work. This parameter is enclosed in square
brackets and is a part of the {\verb|documentclass|} command:
\begin{verbatim}
  \documentclass[STYLE]{acmart}
\end{verbatim}

Journals use one of three template styles. All but three ACM journals
use the {\verb|acmsmall|} template style:
\begin{itemize}
\item {\texttt{acmsmall}}: The default journal template style.
\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
\item {\texttt{acmtog}}: Used by TOG.
\end{itemize}

The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
\begin{itemize}
\item {\texttt{acmconf}}: The default proceedings template style.
\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
\end{itemize}

The template style used for PODC 2023 submissions is {\verb|acmsmall|}.

\subsection{Template Parameters}

In addition to specifying the {\itshape template style} to be used in
formatting your work, there are a number of {\itshape template parameters}
which modify some part of the applied template style. A complete list
of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\texttt{anonymous,review}}: Suitable for a ``double-blind''
  conference submission. Anonymizes the work and includes line
  numbers. Use with the \texttt{\acmSubmissionID} command to print the
  submission's unique ID on each page of the work.
\item{\texttt{authorversion}}: Produces a version of the work suitable
  for posting by the author.
\item{\texttt{screen}}: Produces colored hyperlinks.
\end{itemize}

This document uses the following string as the first command in the
source file:
\begin{verbatim}
\documentclass[acmsmall,nonacm,anonymous]{acmart}
\end{verbatim}

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \verb|\vspace| command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

{\bfseries Your document will be returned to you for revision if
  modifications are discovered.}

\section{Typefaces}

The ``\verb|acmart|'' document class requires the use of the
``Libertine'' typeface family. Your \TeX\ installation should include
this set of packages. Please do not substitute other typefaces. The
``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
as they will override the built-in typeface families.

\section{Title Information}

The title of your work should use capital letters appropriately -
\url{https://capitalizemytitle.com/} has useful rules for
capitalization. Use the {\verb|title|} command to define the title of
your work. If your work has a subtitle, define it with the
{\verb|subtitle|} command.  Do not insert line breaks in your title.

If your title is lengthy, you must define a short version to be used
in the page headers, to prevent overlapping text. The \verb|title|
command has a ``short title'' parameter:
\begin{verbatim}
  \title[short title]{full title}
\end{verbatim}

\section{CCS Concepts and User-Defined Keywords}

Two elements of the ``acmart'' document class provide powerful
taxonomic tools for you to help readers find your work in an online
search.

The ACM Computing Classification System ---
\url{https://www.acm.org/publications/class-2012} --- is a set of
classifiers and concepts that describe the computing
discipline. Authors can select entries from this classification
system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
commands to be included in the \LaTeX\ source.

User-defined keywords are a comma-separated list of words and phrases
of the authors' choosing, providing a more flexible way of describing
the research being presented.

CCS concepts and user-defined keywords are required for for all
articles over two pages in length, and are optional for one- and
two-page articles (or abstracts).

\section{Sectioning Commands}

Your work should use standard \LaTeX\ sectioning commands:
\verb|section|, \verb|subsection|, \verb|subsubsection|, and
\verb|paragraph|. They should be numbered; do not remove the numbering
from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is {\bfseries not allowed.}

\section{Tables}

The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed {\itshape above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and the
table caption.  The contents of the table itself must go in the
\textbf{tabular} environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.  Again,
detailed instructions on \textbf{tabular} material are found in the
\textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \textbf{table*} to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table*}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table*}

Always use midrule to separate table header rows from data rows, and
use it only for this purpose. This enables assistive technologies to
recognise table headers and support their users in navigating tables
more easily.

\section{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections.

\subsection{Inline (In-text) Equations}
A formula that appears in the running text is called an inline or
in-text formula.  It is produced by the \textbf{math} environment,
which can be invoked with the usual
\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
and structures, from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
examples of in-text equations in context. Notice how this equation:
\begin{math}
  \lim_{n\rightarrow \infty}x=0
\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}
A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \textbf{equation}
environment. An unnumbered display equation is produced by the
\textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX\@; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty}x=0
\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  \sum_{i=0}^{\infty} x + 1
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

The ``\verb|figure|'' environment should be used for figures. One or
more images can be placed within a figure. If your figure contains
third-party material, you must clearly identify it as such, as shown
in the example below.
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{sample-franklin}
  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
    Ewing, Inc. [Public domain], via Wikimedia
    Commons. (\url{https://goo.gl/VLCRBB}).}
  \Description{A woman and a girl in white dresses sit in an open car.}
\end{figure}

Your figures should contain a caption which describes the figure to
the reader.

Figure captions are placed {\itshape below} the figure.

Every figure should also have a figure description unless it is purely
decorative. These descriptions convey what’s in the image to someone
who cannot see it. They are also used by search engine crawlers for
indexing images, and when images cannot be loaded.

A figure description must be unformatted plain text less than 2000
characters long (including spaces).  {\bfseries Figure descriptions
  should not repeat the figure caption – their purpose is to capture
  important information that is not already provided in the caption or
  the main text of the paper.} For figures that convey important and
complex new information, a short text description may not be
adequate. More complex alternative descriptions can be placed in an
appendix and referenced in a short figure description. For example,
provide a data table capturing the information in a bar chart, or a
structured list representing a graph.  For additional information
regarding how best to write figure descriptions and why doing this is
so important, please see
\url{https://www.acm.org/publications/taps/describing-figures/}.

\subsection{The ``Teaser Figure''}

A ``teaser figure'' is an image, or set of images in one figure, that
are placed after all author and affiliation information, and before
the body of the article, spanning the page. If you wish to have such a
figure in your article, place the command immediately before the
\verb|\maketitle| command:
\begin{verbatim}
  \begin{teaserfigure}
    \includegraphics[width=\textwidth]{sampleteaser}
    \caption{figure caption}
    \Description{figure description}
  \end{teaserfigure}
\end{verbatim}

\section{Citations and Bibliographies}

The use of \BibTeX\ for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

The bibliography is included in your source document with these two
commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the \BibTeX\ file.

Citations and references are numbered by default. A small number of
ACM publications have citations and references formatted in the
``author year'' style; for these exceptions, please include this
command in the {\bfseries preamble} (before the command
``\verb|\begin{document}|'') of your \LaTeX\ source:
\begin{verbatim}
  \citestyle{acmauthoryear}
\end{verbatim}
